---
title: "Analysis of Quarterly GDP and Key Economic Indicators (V1)"
author: "Kaleb Jordan"
date: "2024-01-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

This study investigates the quarterly United States GDP (Gross Domestic Product) from January 1, 2001 to January 1, 2024 using econometric modeling and time series analysis to forecast GDP over the next two years. Differencing and stationarity tests were applied, followed by ARIMA modeling, residual diagnostics, and forecast evaluation. Key findings suggest that ARIMA(0,2,1) is an applicable model for short-term (2-Year) GDP prediction, with the results highlighting potential implications for economic policymaking.

\newpage

# Introduction

GDP serves as a critical measure of a nation's economic health, providing insights into growth trends, productivity, and overall economic well-being. Accurate GDP forecasting is essential for policymakers and financial institutions to guide economic strategies. This report aims to utilize time series methods to predict future GDP values. Key macroeconomic indicators known to influence GDP are included to provide a comprehensive understanding of economic relationships and future dynamics.

The data set of United States GDP growth consists of 93 observations financial quarters from January 1, 2001-January 1, 2024. Nine other variables are in the data set and can be ignored for the ARIMA only model. *See Appendix for more details.*

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Necessary libraries
library(tidyverse)
library(tseries)
library(forecast)
library(glmnet)
library(lmtest)

# Read GDP csv file
gdp = read_csv("C:/Users/kaleb/Downloads/projs/URES (GDP Growth)/GDP.csv") %>%
  rename(`GDP (In Billions)` = GDP)

# Filtering quarterly data for other indicators as a function
quarterly_filter = function(df) {
  df %>%
    filter((month(observation_date) == 1 & day(observation_date) == 1) |
             (month(observation_date) == 4 & day(observation_date) == 1) |
             (month(observation_date) == 7 & day(observation_date) == 1) |
             (month(observation_date) == 10 & day(observation_date) == 1))
}
# Load other csv files using the quarterly_filter function
unemp = quarterly_filter(read_csv("C:/Users/kaleb/Downloads/projs/URES (GDP Growth)/UNRATE.csv"))
job.open = quarterly_filter(read_csv("C:/Users/kaleb/Downloads/projs/URES (GDP Growth)/JTSJOL.csv"))
lab.part = quarterly_filter(read_csv("C:/Users/kaleb/Downloads/projs/URES (GDP Growth)/CIVPART.csv"))
cpi = quarterly_filter(read_csv("C:/Users/kaleb/Downloads/projs/URES (GDP Growth)/CPALTT01USM657N.csv"))
ppi = quarterly_filter(read_csv("C:/Users/kaleb/Downloads/projs/URES (GDP Growth)/PPIACO.csv"))
stick_cpi = quarterly_filter(read_csv("C:/Users/kaleb/Downloads/projs/URES (GDP Growth)/CORESTICKM159SFRBATL.csv"))
pce = quarterly_filter(read_csv("C:/Users/kaleb/Downloads/projs/URES (GDP Growth)/DPCCRV1Q225SBEA.csv"))
retail = quarterly_filter(read_csv("C:/Users/kaleb/Downloads/projs/URES (GDP Growth)/RSXFS.csv"))
con_sent = quarterly_filter(read_csv("UMCSENT.csv"))

# Combine all variables into one data frame
df = cbind(gdp, unemp$UNRATE, job.open$JTSJOL, lab.part$CIVPART, cpi$CPALTT01USM657N,
            ppi$PPIACO, stick_cpi$CORESTICKM159SFRBATL, pce$DPCCRV1Q225SBEA,
            retail$RSXFS, con_sent$UMCSENT)

# Rename columns
df = df %>%
  rename(Date = observation_date,
         `Unemployment Rate (%)` = `unemp$UNRATE`,
         `Job Openings (In Thousands)` = `job.open$JTSJOL`,
         `Labor Participation (%)` = `lab.part$CIVPART`,
         `Consumer Price Index (Growth)` = `cpi$CPALTT01USM657N`,
         `Producer Price Index (1982=100)` = `ppi$PPIACO`,
         `Sticky CPI` = `stick_cpi$CORESTICKM159SFRBATL`,
         `Personal Consumption Expenditures` = `pce$DPCCRV1Q225SBEA`,
         `Retail Sales (In Millions)` = `retail$RSXFS`,
         `Consumer Sentiment (1966=100)` = `con_sent$UMCSENT`)

# Plot the Quarterly GDP Trend
ggplot(df, aes(x = Date, y = `GDP (In Billions)`)) +
  geom_line(color = "darkgreen", size = 1.2) +
  labs(
    title = "Quarterly GDP Trend",
    x = "Year",
    y = "GDP (In Billions USD)",
    caption = "Source: Federal Reserve Economic Data (FRED)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_x_date(
    breaks = seq(from = min(df$Date), to = max(df$Date), by = "3 years"),
    date_labels = "%Y"  # Formats the labels as year only
  )
```

From the graph, there is a strong linear increase in GDP over most of the observed period, indicating consistent economic growth. Two significant decreases are evident: the first in 2008 and the second in 2020.

-   The decline in 2008 aligns with the global financial crisis and recession during the Bush (43) administration.

-   The drop in 2020 corresponds to the COVID-19 pandemic and economic recession under the Trump administration.

The rising trend suggests non-stationary properties in the data. As a result, stationarity testing will be crucial before conducting time series analysis, with potential transformations (e.g., differencing or log-transformations) required to meet stationarity assumptions.

# Methods

## Data Collection

Quarterly GDP data from 2001–2024 were obtained alongside key macroeconomic variables such as unemployment rate, CPI growth, retail sales, job openings, and labor force participation. Data preprocessing included:

-   Filtering quarterly data.

-   Aggregating and renaming variables for clarity.

-   Converting data into a time series format.

## Stationarity Testing

The Augmented Dickey-Fuller (ADF) test was applied to check stationarity. Differencing was conducted where required to achieve stationarity, critical for time series modeling.

Stationary results of our overall data concluded that the data is non-stationary. A transformation is needed and concluded Differencing is the best solution. *For more information, check Appendix.*

```{r echo=FALSE, message=FALSE, warning=FALSE}
gdp_ts = ts(df$`GDP (In Billions)`, start = c(2001, 1), frequency = 4)
adf.test(gdp_ts)
```

After using differencing to transform the data, we conclude that our data is now stationary but has a p-value just less than 0.05. Consider a second differencing transformation. Here is the plot of differenced GDP values and ADF test results:

```{r echo=FALSE, message=FALSE, warning=FALSE}
gdp_diff = diff(gdp_ts)
plot(gdp_diff, main = "Differenced Quarterly GDP", ylab = "Differenced GDP")
```

Here are the results of the second differencing transformation. We see a significantly lower p-value which ensures stationarity.

```{r echo=FALSE, message=FALSE, warning=FALSE}
gdp_diff2 = diff(gdp_diff)
adf.test(gdp_diff2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
plot(gdp_diff2, main = "Differenced Quarterly GDP", ylab = "Differenced GDP")
```

\newpage

## Model Selection

### ACF/PACF Analysis

Identified potential autoregressive (AR) via PACF graph and moving average (MA) components via ACF graphs for the differenced values.

```{r echo=FALSE, message=FALSE, warning=FALSE}
acf(gdp_diff2, main = "Differenced GDP Values ACF")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
pacf(gdp_diff, main = "Differenced GDP Values PACF")
```

From the ACF graph, we see one significant lag spike which indicates that our ARIMA model will most likely conclude ARIMA(0, 2, 1).

### ARIMA

```{r message=FALSE, warning=FALSE, include=FALSE}
# Set seed for reproducibility
set.seed(88)

# Use auto.arima model to confirm our model
arima_model = auto.arima(gdp_ts, seasonal = FALSE)
summary(arima_model)
```

Selected the optimal ARIMA model based on AIC (Akaike Information Criterion) by running an Auto-ARIMA model.

The Auto-ARIMA model confirms the model concluded by the differencing transformations and ACF/PACF graphs. *For more information on ARIMA models, see Appendix.*

Key insights include:

-   AIC: 1309.07

-   Root Mean Squared Error (RMSE): 308.11

-   Mean Absolute Percentage Error (MAPE): 0.719%

## Forecasting

```{r message=FALSE, warning=FALSE, include=FALSE}
# Forecast 8 quarters or two years of GDP growth for the future
forecast_values = forecast(arima_model, h = 8)
forecast_values

# Plot forecasted values
plot(forecast_values, main = "GDP Forecast", ylab = "GDP", lwd = 3)

forecast_years = c(rep(2024, 4), rep(2025, 4))
forecast_quarters = rep(1:4, times = 2)


forecast_results = data.frame(
  Year = forecast_years,
  Quarter = forecast_quarters,
  Forecasted_GDP = as.numeric(forecast_values$mean),
  Lower_80 = as.numeric(forecast_values$lower[, 1]),
  Upper_80 = as.numeric(forecast_values$upper[, 1]),
  Lower_95 = as.numeric(forecast_values$lower[, 2]),
  Upper_95 = as.numeric(forecast_values$upper[, 2])
)
# Print forecasted values (Same as table in Report)
forecast_results
```

The chosen ARIMA model was used to predict GDP for the next eight quarters (two years). Forecast intervals (80% and 95%) were generated to assess prediction uncertainty.

| Year | Quarter | Point Forecast | 80% Lower Bound | 80% Upper Bound | 95% Lower Bound | 95% Upper Bound |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| 2024 | 2 | 28990.59 | 28589.21 | 29391.97 | 28376.73 | 29604.45 |
| 2024 | 3 | 29357.11 | 28767.05 | 29947.17 | 28454.69 | 30259.53 |
| 2024 | 4 | 29273.63 | 28973.22 | 30474.05 | 28575.97 | 30871.29 |
| 2025 | 1 | 30090.15 | 29191.32 | 30988.98 | 28715.51 | 31464.80 |
| 2025 | 2 | 30456.67 | 29415.31 | 31498.04 | 28864.04 | 32049.30 |
| 2025 | 3 | 30823.19 | 29642.23 | 32004.16 | 29017.06 | 32629.33 |
| 2025 | 4 | 31189.72 | 29870.41 | 32509.02 | 29172.01 | 33207.42 |
| 2026 | 1 | 31556.24 | 30098.83 | 33013.64 | 29327.33 | 33785.15 |

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot forecasted values
plot(forecast_values, main = "GDP Forecast", ylab = "GDP", lwd = 3)
```

# Residual Diagnostics

## Residual Analysis

```{r echo=FALSE, message=FALSE, warning=FALSE}
checkresiduals(arima_model)
```

Overall, the ARIMA(0,2,1) model performs well, as the residuals appear to be normally distributed and show minimal autocorrelation.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# This is a qqplot to visually find if the residuals are normally distributed
qqnorm(arima_model$residuals, main = "Q-Q Plot of Residuals")
qqline(arima_model$residuals, col = "darkgreen", lwd = 2)
```

-   While the residuals are broadly normal, the tail deviations suggest potential issues with capturing rare or extreme events in the data (e.g., anomalies around 2020).

-   The model may still be reliable for forecasting general trends, but it might struggle to handle unexpected shocks.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create a histogram of the model residuals
hist(arima_model$residuals, main = "Residuals Histogram", xlab = "Residuals", breaks = 20, col = "lightgray")
```

-   Residuals have a high frequency close to 0 which indicates good model performance.

-   Large residuals are likely to do with extreme or unexpected events (e.g., anomalies around 2020).

# Future Suggestions

-   **Feature Expansion**: Include more diverse economic indicators and lagged variables for richer modeling.

-   **Multivariate Models**: Apply multivariate time series techniques (e.g., VAR, VECM) to capture dynamic interactions.

-   **Structural Break Analysis**: Investigate potential structural changes during economic events like financial crises.

-   **Real-Time Updates**: Integrate real-time data streams to refine forecasts adaptively.

\newpage

# Appendix

## Abstract

-   All observations are found using data from the following:

    -   Q1: January 1st Annually (2001-2024)

    -   Q2: April 1st Annually (2001-2023)

    -   Q3: July 1st Annually (2001-2023)

    -   Q4: October 1st Annually (2001-2023)

## Introduction

-   Macroeconomic Indicators were not used in ARIMA model but will be included in future ARIMAX models for future versions of this report.

## Methods

### Stationarity Testing

-   When conducting Augmented Dickey-Fuller (ADF) tests, null hypothesis is that the data is non-stationary and needs a transformation. Alternate hypothesis is that the data is stationary and no transformations are needed.

-   When p-value \< 0.05, reject the null hypothesis and conclude the data is stationary.

-   When p-value \> 0.05, accept the null hypothesis and conclude the data is non-stationary and a transformation is needed.

## Model Selection

### ARIMA

-   ARIMA models combine Autoregressive observations, Differencing, and Moving Average terms.

-   Autocorrelation functions (ACF) identify moving average components.

-   Partial Autocorrelation functions (PACF) identify autoregressive components.

### Forecasting

-   Below is the code to understand how the forecasted values were generated:

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Set seed for reproducibility
set.seed(88)

# Use auto.arima model to confirm our model
arima_model = auto.arima(gdp_ts, seasonal = FALSE)
summary(arima_model)

# Forecast 8 quarters or two years of GDP growth for the future
forecast_values = forecast(arima_model, h = 8)
forecast_values

# Plot forecasted values
plot(forecast_values, main = "GDP Forecast", ylab = "GDP", lwd = 3)

forecast_years = c(rep(2024, 4), rep(2025, 4))
forecast_quarters = rep(1:4, times = 2)


forecast_results = data.frame(
  Year = forecast_years,
  Quarter = forecast_quarters,
  Forecasted_GDP = as.numeric(forecast_values$mean),
  Lower_80 = as.numeric(forecast_values$lower[, 1]),
  Upper_80 = as.numeric(forecast_values$upper[, 1]),
  Lower_95 = as.numeric(forecast_values$lower[, 2]),
  Upper_95 = as.numeric(forecast_values$upper[, 2])
)
# Print forecasted values (Same as table in Report)
forecast_results
```

# Source

All data used in this report were from the Federal Reserve Economic Data from the Federal Reserve Bank of St. Louis. All data was tailored to get a specific range of dates (January 1, 2001-January 1, 2024).

-   **Gross Domestic Product (GDP)**\
    Federal Reserve Bank of St. Louis. (2024). *Gross Domestic Product (GDP)* [Dataset]. FRED. Retrieved from <https://fred.stlouisfed.org/series/GDP>

-   **Unemployment Rate (UNRATE)**\
    Federal Reserve Bank of St. Louis. (2024). *Unemployment Rate (UNRATE)* [Dataset]. FRED. Retrieved from <https://fred.stlouisfed.org/series/UNRATE>

-   **Job Openings (JTSJOL)**\
    Federal Reserve Bank of St. Louis. (2024). *Job Openings: Total Nonfarm (JTSJOL)* [Dataset]. FRED. Retrieved from <https://fred.stlouisfed.org/series/JTSJOL>

-   **Labor Force Participation Rate (CIVPART)**\
    Federal Reserve Bank of St. Louis. (2024). *Labor Force Participation Rate (CIVPART)* [Dataset]. FRED. Retrieved from <https://fred.stlouisfed.org/series/CIVPART>

-   **Consumer Price Index (CPI-U)**\
    Federal Reserve Bank of St. Louis. (2024). *Consumer Price Index for All Items (CPALTT01USM657N)* [Dataset]. FRED. Retrieved from <https://fred.stlouisfed.org/series/CPALTT01USM657N>

-   **Producer Price Index (PPI)**\
    Federal Reserve Bank of St. Louis. (2024). *Producer Price Index for All Commodities (PPIACO)* [Dataset]. FRED. Retrieved from <https://fred.stlouisfed.org/series/PPIACO>

-   **Sticky Price CPI**\
    Federal Reserve Bank of St. Louis. (2024). *Sticky Price Consumer Price Index (CORESTICKM159SFRBATL)* [Dataset]. FRED. Retrieved from <https://fred.stlouisfed.org/series/CORESTICKM159SFRBATL>

-   **Personal Consumption Expenditures (PCE)**\
    Federal Reserve Bank of St. Louis. (2024). *Real Personal Consumption Expenditures (DPCCRV1Q225SBEA)* [Dataset]. FRED. Retrieved from <https://fred.stlouisfed.org/series/DPCCRV1Q225SBEA>

-   **Retail Sales (RSXFS)**\
    Federal Reserve Bank of St. Louis. (2024). *Advance Retail Sales: Retail and Food Services, Excluding Motor Vehicles and Parts (RSXFS)* [Dataset]. FRED. Retrieved from <https://fred.stlouisfed.org/series/RSXFS>

-   **University of Michigan Consumer Sentiment Index (UMCSENT)**\
    Federal Reserve Bank of St. Louis. (2024). *University of Michigan: Consumer Sentiment (UMCSENT)* [Dataset]. FRED. Retrieved from <https://fred.stlouisfed.org/series/UMCSENT>
